---
title: "Extended data figure 10"
author: "Marc J Williams"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: kable
    number_sections: yes
    toc: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(grid)
source(here("src/setup.R"))
```

## Background 

In this script we get the scWGS data from the recurrence sample and identify the clones with the highest similarity. We also identify the most similar cells in order to attach them to the tree that was built just using the data from the primary sample.

The below functions do all this and generate some plots.

```{r}
clonecols <- config$clonecols
names(clonecols) <- LETTERS[1:length(clonecols)]

euclidean_distance <- function(p, q) {
  # Check if input vectors have the same length
  if (length(p) != length(q)) {
  stop("Input vectors must have the same length")
  }
  # Calculate the squared differences
  squared_diff <- (p - q)^2
  # Sum the squared differences and take the square root
  distance <- sqrt(sum(squared_diff))
  return(distance)
}


library(data.table)

create_similarity_cells_dt <- function(S2, S1, tree) {
  # Convert inputs to data.table if they aren't already
  setDT(S2)
  setDT(S1)
  
  # Pre-filter S1 to only include cells in the tree
  S1_filtered <- S1[cell_id %in% tree$tip.label]
  
  # Create keys to speed up joins
  setkeyv(S1_filtered, c("chr", "start", "end"))
  setkeyv(S2, c("chr", "start", "end"))
  
  # Function for euclidean distance calculation
  euclidean_distance <- function(a, b) {
    sqrt(sum((a - b)^2, na.rm = TRUE))
  }
  
  # Join only the necessary columns
  similarity_data <- S1_filtered[S2, nomatch=0, allow.cartesian=TRUE,
                      .(chr, start, end, 
                        cell_id.post = i.cell_id, 
                        cell_id.pre = cell_id,
                        copy.post = i.copy, 
                        copy.pre = copy,
                        BAF.post = i.BAF, 
                        BAF.pre = BAF,
                        A.post = i.A, 
                        A.pre = A,
                        B.post = i.B, 
                        B.pre = B)]
  
  # Calculate metrics efficiently
  result <- similarity_data[, .(
    cor = cor(copy.pre, copy.post, use = "complete.obs") + 
         cor(BAF.pre, BAF.post, use = "complete.obs"),
    diff = euclidean_distance(A.post, A.pre) + 
           euclidean_distance(B.post, B.pre)
  ), by = .(cell_id.post, cell_id.pre)]
  
  # Find best match using data.table operations
  similarity_cells <- result[order(diff), .SD[1], by = cell_id.post]
  
  return(similarity_cells)
}

add_new_tips_from_similarity <- function(tree, similarity_df, branch_length_method = "scaled") {
  # Validate that all pre-cells exist in the tree
  missing_tips <- similarity_df$cell_id.pre[!similarity_df$cell_id.pre %in% tree$tip.label]
  if (length(missing_tips) > 0) {
    warning("Some reference cells are not in the tree: ", paste(unique(missing_tips), collapse=", "))
    similarity_df <- similarity_df %>% filter(cell_id.pre %in% tree$tip.label)
  }
  
  # Remove duplicate post cells (if any), keeping only the most similar match
  similarity_df <- similarity_df %>%
    arrange(diff) %>%  # Sort by difference (assuming smaller is more similar)
    distinct(cell_id.post, .keep_all = TRUE)
  
  # Make a copy of the original tree
  modified_tree <- tree
  
  # Process each new tip
  for (i in 1:nrow(similarity_df)) {
    new_tip <- similarity_df$cell_id.post[i]
    ref_tip <- similarity_df$cell_id.pre[i]
    
    # Skip if new tip is already in the tree
    if (new_tip %in% modified_tree$tip.label) {
      message("Tip ", new_tip, " already exists in the tree. Skipping.")
      next
    }
    
    # Calculate branch length
    if (branch_length_method == "scaled") {
      # Scale branch length based on similarity/difference
      # Lower diff means more similar, so we want shorter branches
      branch_length <- similarity_df$diff[i] / 100  # Scale factor, adjust as needed
    } else if (branch_length_method == "fixed") {
      # Use a fixed small branch length
      branch_length <- 0.01
    } else if (branch_length_method == "proportional") {
      # Proportional to the existing branch lengths
      closest_tip_index <- which(modified_tree$tip.label == ref_tip)
      edge_index <- which(modified_tree$edge[,2] == closest_tip_index)
      parent_branch_length <- modified_tree$edge.length[edge_index]
      branch_length <- parent_branch_length * 0.5  # Adjust as needed
    }
    
    # Create a new small tree with the new tip
    new_tip_tree <- stree(1)
    new_tip_tree$tip.label <- new_tip
    new_tip_tree$edge.length <- branch_length
    
    # Find where to attach in the main tree (node connecting to ref_tip)
    ref_tip_index <- which(modified_tree$tip.label == ref_tip)
    edge_to_ref <- which(modified_tree$edge[,2] == ref_tip_index)
    parent_node <- modified_tree$edge[edge_to_ref, 1]
    
    # Bind the new tip to the parent node of the reference tip
    modified_tree <- bind.tree(modified_tree, new_tip_tree, where = parent_node)
    
    message("Added tip ", new_tip, " near ", ref_tip)
  }
  
  return(modified_tree)
}

get_plots <- function(mypatient){

  #cn10kb <- read_allfiles(pattern = "_clone_copynumber_10kb.csv", sep = "_clone_copynumber_10kb.csv", patients = c(mypatient))
  clones <- read_allfiles(pattern = "_clones.csv", sep = "_clones.csv", patients = c(mypatient))
  if (mypatient == "026"){
    clones <- clones %>% mutate(clone_id = ifelse(clone_id == "C", "B", clone_id))
  }
  if (mypatient == "046"){
    clones <- clones %>% filter(clone_id != "C")
  }
  hscn <- read_allfiles(pattern = "_hscn.csv.gz", sep = "_hscn.csv.gz", patients = c(mypatient)) %>% 
    filter(cell_id != "SPECTRUM-OV-026_S2_INFRACOLIC_OMENTUM-130382A-R43-C28") #remove as not a cancer cell
  metrics <- read_allfiles(pattern = "_qc.csv.gz", sep = "_qc.csv.gz", patients = c(mypatient))
  cells_to_keep <- metrics %>% filter(fracLOH_17 > 0.9) %>% pull(cell_id)
  hscn <- filter(hscn, cell_id %in% cells_to_keep)
  S1 <- hscn %>% filter(str_detect(cell_id, "S1"))
  S1cons <- S1 %>% consensuscopynumber(., cl = clones %>% filter(clone_id != "0"))
  S2 <- hscn %>% filter(str_detect(cell_id, "S2"))
  S2cons <- S2 %>% consensuscopynumber()
  
  similarity <- inner_join(S2, S1cons, by = c("chr", "start", "end"), suffix = c(".post", ".pre")) %>% 
    group_by(cell_id.post,  cell_id.pre) %>% 
    summarize(cor = cor(copy.pre, copy.post, use = "complete.obs") + cor(BAF.pre, BAF.post, use = "complete.obs"),
              diff = euclidean_distance(A.post, A.pre) + euclidean_distance(B.post, B.pre)) %>% 
    #arrange(desc(cor)) %>% 
    arrange(diff) %>% 
    filter(row_number() == 1)  %>% 
    rename(cell_id = cell_id.post, clone_id = cell_id.pre)
  #similarity_cells <- similarity
  
  chroms <- c(paste0(1:11), "13", "15", "17", "20", "X")
  
  p1 <- plotHeatmap(S2,# %>% filter(chr == "22"), #clone_pal = clonecols,
              show_library_label = F,
              annofontsize = 5,
              linkheight = 1.5,
              chrlabels = chroms,
              anno_width = 0.2,
              clone_pal = clonecols,
              clusters = similarity,
              show_heatmap_legend = FALSE,
              show_legend = FALSE,
              tree = NULL,
              plottree = F,
              reorderclusters = T)
  p1_gg <- grid.grabExpr(ComplexHeatmap::draw(p1), 
                         width = 89 * 0.039, height = 2)
  
  p2 <- plotHeatmap(S2,# %>% filter(chr == "22"), #clone_pal = clonecols,
              show_library_label = F,
              clone_pal = clonecols,
              show_legend = FALSE,
              annofontsize = 5,
              linkheight = 1.5,
              anno_width = 0.2,
              chrlabels = chroms,
              clusters = similarity,
              show_heatmap_legend = FALSE,
              plotcol = "state_phase",
              tree = NULL,
              plottree = F,
              reorderclusters = T)
  p2_gg <- grid.grabExpr(ComplexHeatmap::draw(p2), 
                         width = 89 * 0.039, height = 2)
  
  
  p3 <- plotHeatmap(S1cons,# %>% filter(chr == "22"), #clone_pal = clonecols,
              show_library_label = F,
              clone_pal = clonecols,
              show_heatmap_legend = FALSE,
              show_legend = FALSE,
              annofontsize = 5,
              linkheight = 1.5,
              anno_width = 0.2,
              chrlabels = chroms,
              clusters = S1cons %>% distinct(cell_id) %>% mutate(clone_id = cell_id),
              tree = NULL,
              plottree = F,
              reorderclusters = T)
  p3_gg <- grid.grabExpr(ComplexHeatmap::draw(p3), 
                         width = 89 * 0.039, height = 2)
  
  p4 <- plotHeatmap(S1cons,# %>% filter(chr == "22"), #clone_pal = clonecols,
              show_library_label = F,
              show_legend = FALSE,
              show_heatmap_legend = FALSE,
              clone_pal = clonecols,
              plotcol = "state_phase",
              annofontsize = 5,
              linkheight = 1.5,
              anno_width = 0.2,
              chrlabels = chroms,
              clusters = S1cons %>% distinct(cell_id) %>% mutate(clone_id = cell_id),
              tree = NULL,
              plottree = F,
              reorderclusters = T)
  p4_gg <- grid.grabExpr(ComplexHeatmap::draw(p4), 
                         width = 89 * 0.039, height = 2)
  
  glist_all <- plot_grid(p3_gg,
                      p1_gg,
                      p4_gg,
                      p2_gg, ncol = 4)
  
  glist_tot <- plot_grid(p3_gg,
                    p1_gg, ncol = 2)
  
  tree <- ape::read.tree(find_allfiles("newick", patient = mypatient))
  if (mypatient != "046"){
    similarity_cells <- create_similarity_cells_dt(S2, S1, tree)
    tree <- add_new_tips_from_similarity(tree, similarity_cells, "scaled")
    clones <- clones %>% select(cell_id, clone_id) %>% 
      bind_rows(similarity %>% select(cell_id, clone_id))
  } else{
    similarity_cells <- NULL
  }

  annodf <- data.frame(cell_id = unlist(tree$tip.label %>% unique)) %>% 
    na.omit() %>% 
    left_join(clones %>% select(cell_id, clone_id)) %>% 
    mutate(surgery = ifelse(str_detect(cell_id, "S1"), "S1", "S2")) %>% 
    replace_na(list(clone_id = "0")) %>% 
    left_join(hscn %>% group_by(cell_id) %>% summarize(ploidy = paste0(round(mean(state))))) %>% 
    filter(cell_id %in% unique(hscn$cell_id)) %>% 
    filter(clone_id != "0") %>% 
    as.data.frame()
  
  p1_tree <- plotHeatmap(hscn,# %>% filter(chr == "22"), #clone_pal = clonecols,
            show_library_label = F,
            annotations = annodf,
            tree = tree,
            annofontsize = 5,
            linkheight = 1.5,
            chrlabels = chroms,
            anno_width = 0.3,
            #clone_pal = clonecols,
            clusters = annodf %>% mutate(clone_id = "0"),
            show_heatmap_legend = FALSE,
            show_legend = FALSE,
            plottree = T,
            tree_width = 3,
            reorderclusters = T)
  p1_tree_gg <- grid.grabExpr(ComplexHeatmap::draw(p1_tree), 
                         width = 89 * 0.039, height = 2)
  
  p2_tree <- plotHeatmap(hscn,# %>% filter(chr == "22"), #clone_pal = clonecols,
            show_library_label = F,
            annotations = annodf,
            tree = tree,
            annofontsize = 5,
            linkheight = 1.5,
            chrlabels = chroms,
            anno_width = 0.1,
            plotcol = "state_phase",
            #clone_pal = clonecols,
            clusters = annodf %>% mutate(clone_id = "0"),
            show_heatmap_legend = FALSE,
            show_legend = TRUE,
            plottree = F,
            tree_width = 1.6,
            reorderclusters = T)
  p2_tree_gg <- grid.grabExpr(ComplexHeatmap::draw(p2_tree), 
                         width = 89 * 0.039, height = 2)
  
  return(list(glist_all = glist_all,
              glist_tot = glist_tot,
              similarity = similarity, 
              similarity_cells = similarity_cells, 
              S1 = S1,
              S1cons = S1cons,
              S2 = S2,
              S2cons = S2cons,
              clones = clones,
              #cn10kb = cn10kb,
              metrics = metrics,
              p1_tree_gg = p1_tree_gg,
              p2_tree_gg = p2_tree_gg
              ))
}

```


## Generate plots for 026, 139 and 046

```{r}
x_026 <- get_plots("026")
x_139 <- get_plots("139")
x_046 <- get_plots("046")
```

## OV-026 WGD

Generate copy number profiles from whole genome doubled cell found in clone C and consensus copy number of all recurrence cells.

```{r}
g1 <- plotCNprofile(x_026$S1, 
              cellid = "SPECTRUM-OV-026_S1_LEFT_OVARY-128742A-R61-C49", 
              legend.position = "none", font_size = 6, line_size = 0.25, pointsize = 0.2) +
  ggtitle("026 Pre-treatment WGD cell (clone C)")  +
  theme(plot.title = element_text(face = "plain"))

g2 <- plotCNprofile(x_026$S2 %>% consensuscopynumber(), 
              #cellid = "SPECTRUM-OV-026_S1_LEFT_OVARY-128742A-R61-C49", 
              legend.position = "none", font_size = 6, line_size = 0.25, pointsize = 0.2) +
  ggtitle("026 recurrence consensus copynumber")  +
  theme(plot.title = element_text(face = "plain"))

```

## Generate final figures

```{r, fig.height = 6, fig.width = 5}
plotlist <- readRDS(here("data/muller_plots.rds"))

top1 <- plot_grid(plotlist[["026"]],plotlist[["139"]],plotlist[["046"]], ncol = 3, 
          labels = c("a 026", "b 139", "c 046"), label_size = 8)

bottom1 <- plot_grid(plot_grid(NULL, x_026$glist_tot, ncol = 2, rel_widths = c(1,3)),
                     plot_grid(NULL, x_139$glist_tot, ncol = 2, rel_widths = c(1,3)),
                     plot_grid(NULL, x_046$glist_tot, ncol = 2, rel_widths = c(1,3)),
                     ncol = 1, 
                    labels = c("a 026", "b 139", "c 046"), label_size = 7) 


plot_grid(bottom1,
          g1,
          g2, ncol = 1, rel_heights = c(3,0.7,0.7), labels = c("", "d", "e"), label_size = 7) %>% 
    save_plot(.,
            filename = here("Figures/EDF10_recurrence.pdf"), 
            base_width = config$default_plot_size$width,
            base_height = 6)


plot_grid(bottom1,
          g1,
          g2, ncol = 1, rel_heights = c(3,0.7,0.7), labels = c("", "d", "e"), label_size = 7) 
```


```{r}
myplot <- plot_grid(x_026$p1_tree_gg,
          x_139$p1_tree_gg,
          x_046$p1_tree_gg, ncol = 3)
save_plot(myplot,
      filename = here("Figures/other/recurrence_trees.pdf"), 
      base_width = config$default_plot_size$width, nrow = 2,
      base_height = 0.7)

```