---
title: "Figure 4"
author: "Marc J Williams"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: kable
    number_sections: yes
    toc: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup

```{r}
library(here)
source(here("src/setup.R"))
source(here("src/clone_trees.R"))
source(here("src/muller_plots.R"))
clonecols <- config$clonecols
names(clonecols) <- LETTERS[1:length(clonecols)]
```

## Read in data

```{r}
ctdna_patients <- config$patients
ctdna_patients <- unlist(lapply(ctdna_patients, function(x) str_remove(x, "SPECTRUM-OV-")))

meta_long <- readxl::read_xlsx(config$tables, sheet = "S11 - clinical data") %>% 
  mutate(time_from_dx = as.numeric(time_from_dx)) %>% 
  mutate(Final_tp = as.numeric(Final_tp))
recurrence <- readxl::read_xlsx(config$tables, sheet = "S12 - recurrence data") %>% 
  mutate(date = as.numeric(date))
ca125 <- readxl::read_xlsx(config$tables, sheet = "S13 - CA125")
surgeries <- readxl::read_xlsx(config$tables, sheet = "S14 - surgeries") %>% 
  mutate(surgery_date = as.numeric(surgery_date))
svdf_clone <- readxl::read_xlsx(config$tables, sheet = "S3 - Clone frequencies (SVs)") %>% 
  mutate(nclones = 1) %>% 
  left_join(purity %>% select(patient, timepoint, tumorfraction_ClonalSV))

svdf_clone <- svdf_clone %>% 
  mutate(clone_frequency = ifelse(tumorfraction_ClonalSV < 0.0001, 0, clone_frequency)) %>% 
  mutate(clone_frequency_normalized = ifelse(tumorfraction_ClonalSV < 0.0001, 0, clone_frequency_normalized))

treatcols <- c("Alive" = "yellowgreen", 
               "Deceased" = "black", 
               "Anti Angiogenic" = "mediumpurple3", 
               "Chemotherapy" = "darkorange1",
               "Immunotherapy" = "goldenrod2", 
               "MIRV" = "mediumblue", 
               "PARPi" = "mediumseagreen", 
               "T-DXd" = "deepskyblue3")
```


## Plot clone trees

```{r}
mypts <- config$patients_timeline
mypts <- unlist(lapply(mypts, function(x) str_remove(x, "SPECTRUM-OV-")))
mytreeplots <- lapply(mypts, clone_tree_plot)
names(mytreeplots) <- mypts

plot_grid(plotlist = mytreeplots, ncol = 3, labels = names(mytreeplots), label_size = 7)
```

## Plot muller plots

Note that for PARPi the records of administration are incomplete, because patients do not need to come to the hospital to recieve treatment. Therefore it looks like there are gaps in the PARPi treatment but actually patients recieve PARPi continuously (with occasional suspension due to illness or other treatments). In the final figures we fill in these gaps manual having confirmed that the start and end dates are correct with clinical colleagues.

For the muller plots, we also force ctDNA and clone frequencies to be 0 if there are long periods of time without cfDNA timepoints and CA-125 levels suggest the absence of detectable tumor DNA.

```{r}

plot_timeline <- function(pts, muller_smooth_cutoff = 1.0, zerotimepoints = NULL, type = "muller2", ...){
  plot_timeline_newv(pts,  
                     recurrence %>% filter(recurrence_number == 1),
                     plotTP53 = F, 
                     svdf_clone, 
                     surgeries,  
                     ca125 = ca125, 
                     purity = purity, 
                     meta_long = meta_long, 
                     treatcols = treatcols, 
                     gap = gap1, 
                     gap2 = gap2, 
                     line_size = 0.1, 
                     type = type, 
                     zerotimepoints = zerotimepoints,
                     muller_smooth_cutoff = muller_smooth_cutoff,
                    ...)
}

gap1 <- -0.1
gap2 <- -0.1


plotlist <- list()
plotlist[["002"]] <- plot_timeline("002")
plotlist[["004"]] <- plot_timeline("004")
plotlist[["006"]] <- plot_timeline("006", zerotimepoints = c(150, 375))
plotlist[["009"]] <- plot_timeline("009", zerotimepoints = c(365, 1095))
plotlist[["014"]] <- plot_timeline("014", zerotimepoints = c(200, 600), overwritelastT = 750)
plotlist[["022"]] <- plot_timeline("022", zerotimepoints = c(100, 530))
plotlist[["026"]] <- plot_timeline("026", zerotimepoints = c(100, 1500), overwritelastT = 2000)
plotlist[["037"]] <- plot_timeline("037", zerotimepoints = c(100, 365))
plotlist[["044"]] <- plot_timeline("044",  zerotimepoints = c(150, 500), overwritelastT = T)
plotlist[["045"]] <- plot_timeline("045")
plotlist[["046"]] <- plot_timeline("046")
plotlist[["051"]] <- plot_timeline("051")
plotlist[["075"]] <- plot_timeline("075", zerotimepoints = c(200, 380), overwritelastT = 500)
plotlist[["081"]] <- plot_timeline("081", zerotimepoints = c(110, 400))
plotlist[["083"]] <- plot_timeline("083", zerotimepoints = c(110, 1100))
plotlist[["107"]] <- plot_timeline("107", zerotimepoints = c(100, 530))
plotlist[["129"]] <- plot_timeline("129", zerotimepoints = c(100, 530))
plotlist[["139"]] <- plot_timeline("139", zerotimepoints = c(100, 530))

for (pts in mypts){
  save_plot(here(glue::glue("Figures/timelines/{pts}_timeline.png")), plotlist[[pts]], base_height = 1.9, base_width = 2/3 * 89 * 0.039)
}

cowplot::plot_grid(plotlist = plotlist, 
                   ncol = 5,
                   labels = names(plotlist),
                   label_size = 8) %>% 
  save_plot(here("Figures/timelines/mullerplots_wide.pdf"), 
          ., base_width = 2.9 * 89 * 0.039, base_height =8)

saveRDS(plotlist, file = here("data/muller_plots.rds"))
```

## Genetic alterations

Plot some of the clone specific genetic alterations.

```{r}
source(here("src/utils.R"))
geneplots <- list()

mydnageneplot <- function(mygene, myylab = NULL, comps = list(c("A", "B"))){
  symnum.args <- list(cutpoints = c( 0.00, 0.001, 0.01, 0.05, 1), symbols = c("***", "**", "*", "ns"))
  cp <- config$clonecols
  names(cp) <- LETTERS[1:length(cp)]
  
  if (is.null(myylab)){
    myylab <- paste0(mygene, " CN")
  }
  genedf <- gene_copynumber(cn, mygene) %>% 
  left_join(clones)
  
  genedf %>% 
      filter(clone_id %in% comps[[1]]) %>% 
      t.test(copy ~ clone_id, data = .) %>% 
      print()

  gdna <- genedf %>% 
      filter(clone_id != "0") %>% 
      ggplot(aes(x = clone_id, y = copy, fill = clone_id, col = clone_id)) +
      geom_violin(scale = "width",  alpha = 0.4, size = 0.025) +
      geom_boxplot(outlier.colour = NA, color = "white", alpha = 0.8, size = 0.1, width = 0.28) +
      #geom_jitter(height = 0, width = 0.1, size = 0.3, alpha = 0.8) +
      xlab("") +
      scale_fill_manual(values = cp) +
      scale_color_manual(values = cp) +
      theme(legend.position = "none") +
      ylab(myylab) +
      ggpubr::stat_compare_means(label.y.npc = 0.9,
                                 method = "t.test",
                                 aes(label = paste0(..p.signif..)), 
                                 size = 2.5, symnum.args = symnum.args,
                                 comparisons = comps)
  return(gdna)
}

mysample <- "044"
cn_10kb <- read_allfiles(pattern = "_clone_copynumber_10kb.csv", mysample) %>% 
  filter(mappability > 0.99) %>% 
  filter(!clone_id %in% c("0", "Pseudobulk"))
clones <- read_allfiles(pattern = "clones.csv", mysample)
cn <- read_allfiles(pattern = "hscn.csv.gz", mysample)
clones <- clones
g044_1 <- mydnageneplot("ERBB2", comps = list(c("B", "E")))

g044_cn <- plotCNprofile(cn_10kb[,cell_id := clone_id], chrstart = 25, chrend = 45,
                         tickwidth = 10,
              cellid = "B", 
              pointsize = 0.4,
              chrfilt = c("17"),
              SV = svs_md %>% filter(target_idx == "17_37721055") %>% as.data.frame(),
              ideogram = F,
              maxCN = 30,
              font_size = 7,
              line_size = 0.25,
              #annotateregions = gene_locations$hg19 %>% filter(ensembl_gene_symbol == "ERBB2"),
              y_axis_trans = "squashy"
              )

recurrence_patient_044 <- recurrence %>% filter(str_detect(patient_id, "044")) %>% filter(row_number() == 1)
g044_tra <- svs %>% 
  filter(target_idx == "target_17_37721055") %>% 
  filter(bamtype == "uncollapsed") %>% 
  select(timepoint, alt_counts, alt_seq_reads, days) %>% 
  arrange(timepoint) %>% 
  ggplot(aes(x = days, y = alt_seq_reads)) +
  geom_line(size = 0.5, col = "grey70") +
  geom_point(size = 0.4) +
  geom_rect(data = recurrence_patient_044,
                                 aes(xmin = date, x = NULL, y = NULL, fill = NULL, col = NULL), 
                                 xmax = Inf, ymin = -Inf, ymax = Inf, 
                                 fill = "grey60", color = NA, alpha = .2) +
  scale_x_continuous(breaks = scales::breaks_width(365), guide = guide_axis(angle = 45)) +
  scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10), breaks = c(0, 10, 30, 100, 500)) +
  xlab("Days") +
  ylab("# reads")

geneplots[[paste0(mysample)]] <- mydnageneplot("ERBB2", comps = list(c("B", "E")))

geneplots[[paste0(mysample, "v2")]] <- plot_grid(g044_1 + ylab("ERBB2"), 
                    NULL,
                    g044_cn, 
                    NULL,
                    ncol = 1, 
                    align = "v", 
                    axis = "lr",
                    rel_heights = c(1, -0.1, 1, -0.3))

mysample <- "107"
clones <- read_allfiles(pattern = "clones.csv", mysample)
cn <- read_allfiles(pattern = "hscn.csv.gz", mysample)
g107_3a <- mydnageneplot("NOTCH3", comps = list(c("A", "D")))
g107_3b <- mydnageneplot("CCNE1", comps = list(c("A", "D")))
geneplots[[mysample]] <- plot_grid(g107_3a + removexaxis + ylab("NOTCH3"), 
                    NULL,
                    g107_3b + ylab("CCNE1"), 
                    NULL,
                    ncol = 1, 
                    align = "v", 
                    axis = "lr",
                    rel_heights = c(1, -0.1, 1.3, -0.3))

mysample <- "045"
clones <- read_allfiles(pattern = "clones.csv", mysample)
cn <- read_allfiles(pattern = "hscn.csv.gz", mysample)
g045_3 <- mydnageneplot("RAB25", comps = list(c("A", "D")))
g045_3a <- mydnageneplot("RAB25", comps = list(c("A", "D")))
g045_3b <- mydnageneplot("CCNE1", comps = list(c("A", "D")))

geneplots[[mysample]] <- plot_grid(g045_3a + removexaxis + ylab("RAB25"), 
                    NULL,
                    g045_3b + ylab("CCNE1"), 
                    NULL,
                    ncol = 1, 
                    align = "v", 
                    axis = "tblr",
                    rel_heights = c(1, -0.1, 1.3, -0.3))

mysample <- "009"
library(ggtranscript)
recurrence_patient <- recurrence %>% filter(str_detect(patient_id, "009"))
gbrcarev <- svdf_clone %>% 
  filter(str_detect(patient, "009")) %>% 
  distinct(timepoint, days) %>% 
  mutate(nreads = c(rep(0, 10), 1,2,6)) %>% 
  ggplot(aes(x = days, y = nreads)) +
  geom_line(size = 0.5, col = "grey70") +
  geom_point(size = 0.4) +
  geom_rect(data = recurrence_patient,
                                 aes(xmin = date, x = NULL, y = NULL, fill = NULL, col = NULL), 
                                 xmax = Inf, ymin = -Inf, ymax = Inf, 
                                 fill = "grey60", color = NA, alpha = .2) +
  scale_x_continuous(breaks = scales::breaks_width(365), guide = guide_axis(angle = 45)) +
  scale_y_continuous(limits = c(-0.5, 7)) +
  xlab("Days") +
  ylab("# reads")

brca1exons <- fread(here("data/BRCA1-exons.csv"))
brca1exons <- brca1exons %>% 
  dplyr::rename(seqnames = chr) %>% 
  mutate(strand = "+", gene_name = "BRCA1", type = "exon") %>% 
  mutate(transcript_name = "BRCA1") %>% 
  filter(exon_number %in% c(10, 9))

brca1del <- data.frame(seqnames = "17", start = 41246495, end = 41247865) %>% 
  mutate(transcript_name = "BRCA1")

example_mutation <- dplyr::tibble(
  transcript_name = "BRCA1", 
  position = 41246508
)

brca1plot <- brca1exons %>%
    ggplot(aes(
        xstart = start,
        xend = end,
        y = transcript_name
    )) +
    geom_range(fill = "deepskyblue3", col = "deepskyblue3") +
    geom_intron(
        data = to_intron(brca1exons, "transcript_name"),
        strand = "-"
    ) +
  geom_junction(
    data = brca1del,
    junction.y.max = 0.5
  ) +
  geom_vline(
    data = example_mutation, 
    aes(
      xintercept = position, 
      xstart = NULL,
      xend = NULL
      ), 
    linetype = 2,
    colour = "red"
  ) +
  theme_void()

brca1plot <- plot_grid(NULL, brca1plot, NULL, ncol = 1, rel_heights = c(0.2,1,0.2))

geneplots[[mysample]] <- plot_grid(plot_grid(NULL, brca1plot, NULL,  
                    ncol = 3, 
                    rel_widths = c(0.4, 1, 0.2), 
                    labels = c(""), label_size = 8), 
                    gbrcarev, ncol = 1, rel_heights = c(0.6, 1))
```

Now plot alterations in patients included in supplementary figures.

```{r}
geneplots2 <- list()
geneplots2_10kb <- list()

mysample <- "046"
clones <- read_allfiles(pattern = "clones.csv", mysample)
cn <- read_allfiles(pattern = "hscn.csv.gz", mysample)
cn_10kb <- read_allfiles(pattern = "_clone_copynumber_10kb.csv", mysample) %>% 
  filter(mappability > 0.99) %>% 
  filter(!clone_id %in% c("0", "Pseudobulk")) %>% 
  mutate(cell_id = clone_id)
geneplots2[["046-FGFR3"]] <- mydnageneplot("FGFR3",comps = list(c("A", "B")))
geneplots2[["046-HLA"]] <- mydnageneplot("HLA-A", comps = list(c("A", "B")))
  
g1_046 <- plotCNprofile(cn_10kb,chrstart = 0, chrend = 50, tickwidth = 10, ideogram = F, maxCN = 8,
              cellid = "A", 
              pointsize = 0.01, raster = T,
              annotateregions = gene_locations$hg19 %>% filter(ensembl_gene_symbol == "FGFR3"),
              chrfilt = "4", 
              legend.position = "none",
              font_size = 8, line_size = 0.25)
g2_046 <- plotCNprofile(cn_10kb,chrstart = 0, chrend = 50, tickwidth = 10, ideogram = F, maxCN = 8,
              cellid = "B", 
              pointsize = 0.01, raster = T,
              annotateregions = gene_locations$hg19 %>% filter(ensembl_gene_symbol == "FGFR3"),
              chrfilt = "4", 
              legend.position = "none",
              font_size = 8, line_size = 0.25)
plot_grid(plot_grid(g1_046 + removexaxis + ylab("Clone A CN"), 
          g2_046 + ylab("Clone B CN"), ncol = 1),
          geneplots2[["046-FGFR3"]])

mysample <- "022"
clones <- read_allfiles(pattern = "clones.csv", mysample)
cn <- read_allfiles(pattern = "hscn.csv.gz", mysample)
cn_10kb <- read_allfiles(pattern = "_clone_copynumber_10kb.csv", mysample) %>% 
  filter(mappability > 0.99) %>% 
  filter(!clone_id %in% c("0", "Pseudobulk")) %>% 
  mutate(cell_id = clone_id)
geneplots2[["022-MYC"]] <- mydnageneplot("MYC", comps = list(c("A", "E")))

g1_022 <- plotCNprofile(cn_10kb,#chrstart = 0, chrend = 50, tickwidth = 10, ideogram = F, maxCN = 8,
              cellid = "A", 
              pointsize = 0.01, raster = T,
              annotateregions = gene_locations$hg19 %>% filter(ensembl_gene_symbol == "MYC"),
              chrfilt = "8", 
              legend.position = "none",
              font_size = 8, line_size = 0.25)
g2_022 <- plotCNprofile(cn_10kb,#chrstart = 0, chrend = 50, tickwidth = 10, ideogram = F, maxCN = 8,
              cellid = "E", 
              pointsize = 0.01, raster = T,
              annotateregions = gene_locations$hg19 %>% filter(ensembl_gene_symbol == "MYC"),
              chrfilt = "8", 
              legend.position = "none",
              font_size = 8, line_size = 0.25)

g4 <- plot_grid(NULL,
          geneplots2[["046-FGFR3"]], 
          NULL,
          geneplots2[["022-MYC"]],
          NULL, nrow = 1, rel_widths = c(0.1,1,0.4, 1, 0.1))


plot_grid(plot_grid(plot_grid(g1_046 + removexaxis + ylab("Clone A CN"), 
          g2_046 + ylab("Clone B CN"), ncol = 1),
          plot_grid(plot_grid(g1_022 + removexaxis + ylab("Clone A CN"), 
          g2_022 + ylab("Clone E CN"), ncol = 1)), 
          labels = c("046", "022"), label_size = 7),
          g4, ncol = 1, rel_heights = c(1, 0.6)) %>% 
  save_plot(plot = ., filename = here("Figures/other/022_046_hlamps.pdf"), base_height = 3, base_width = config$default_plot_size$width / 2)



mysample <- "083"
clones <- read_allfiles(pattern = "clones.csv", mysample)
cn <- read_allfiles(pattern = "hscn.csv.gz", mysample)
cn_10kb <- read_allfiles(pattern = "_clone_copynumber_10kb.csv", mysample) %>% 
  filter(mappability > 0.99) %>% 
  filter(!clone_id %in% c("0", "Pseudobulk")) %>% 
  mutate(cell_id = clone_id)

g1_083 <- plotCNprofile(cn_10kb,#chrstart = 0, chrend = 50, tickwidth = 10, ideogram = F, maxCN = 8,
              cellid = "A", 
              pointsize = 0.01, raster = T,
              #annotateregions = gene_locations$hg19 %>% filter(ensembl_gene_symbol == "MYC"),
              chrfilt = c("6", "8"), 
              maxCN = 25,
              y_axis_trans = "squashy",
              legend.position = "none",
              font_size = 8, line_size = 0.25)
g2_083 <- plotCNprofile(cn_10kb,#chrstart = 0, chrend = 50, tickwidth = 10, ideogram = F, maxCN = 8,
              cellid = "B", 
              pointsize = 0.01, raster = T,
              #annotateregions = gene_locations$hg19 %>% filter(ensembl_gene_symbol == "MYC"),
              chrfilt = c("6", "8"), 
               maxCN = 25,
              y_axis_trans = "squashy",
              legend.position = "none",
              font_size = 8, line_size = 0.25)

plot_grid(plot_grid(g1_083 + removexaxis + ylab("Clone A CN"), 
          g2_083 + ylab("Clone B CN"), ncol = 1))  %>% 
  save_plot(plot = ., filename = here("Figures/other/083_chromothripsis.pdf"), base_height = 3, base_width = config$default_plot_size$width / 4)


```

```{r}
plotlistmain <- list()
i <- 1
for (pts in "044"){
  print(pts)
  treeplot <- plot_grid(ggplot() + theme_cowplot(font_size = 7) +
                          removexaxis + removeyaxis + ggtitle(paste0(letters[i], "   ", pts)), 
                        mytreeplots[[pts]], NULL, NULL,
                        ncol = 2, 
                        rel_widths = c(1,1),
                        rel_heights = c(1.3,1))
  plotlistmain[[pts]] <- cowplot::plot_grid(treeplot, 
                                           plotlist[[pts]], 
                                           plot_grid(NULL, geneplots[[pts]], NULL, 
                                                     ncol = 1, rel_heights = c(0.3, 1, 0.3)),
                                           ncol = 3,
                                           rel_widths = c(1.2, 1.8, 0.6))
  i <- i + 1
}

main_fig_pts <- c("009", "107", "045", "026", "139", "046")
i <- 1
for (pts in main_fig_pts){
  print(pts)
  treeplot <- plot_grid(ggplot() + theme_cowplot(font_size = 7) +
                          removexaxis + removeyaxis + ggtitle(paste0(letters[i], "   ", pts)), 
                        mytreeplots[[pts]], NULL, NULL,
                        ncol = 2, 
                        rel_widths = c(1,1),
                        rel_heights = c(1.3,1))
  plotlistmain[[pts]] <- cowplot::plot_grid(treeplot, 
                                           plotlist[[pts]], 
                                           ncol = 2,
                                           rel_widths = c(0.5, 1))
  i <- i + 1
}

```


```{r}
for_leg <- plot_timeline("009", zerotimepoints = c(365, 1095), plotlist = T)
leg <- get_legend(for_leg$gclone)

gmuller_main <- plot_grid(plotlistmain[["044"]],
          NULL,
          plot_grid(plotlist = plotlistmain[c("009", "107", "045")], ncol = 3),
          plot_grid(gbrcarev, geneplots[["107"]], geneplots[["045"]], NULL, rel_widths = c(1.33, 0.89, 0.89, 0.89), nrow = 1),
          plot_grid(plotlist = plotlistmain[c("026", "139", "046")], ncol = 3),
          NULL,
          ncol = 1, rel_heights = c(1,0.6,0.8,0.4,0.8, 0.25))

gmuller_main %>% 
  save_plot(here("Figures/Fig4_mullerplots.pdf"), 
          ., base_width = 2 * 89 * 0.039, base_height =8)
```


```{r}

mysample <- "044"
cn_10kb <- read_allfiles(pattern = "_clone_copynumber_10kb.csv", mysample) %>% 
  mutate(cell_id = clone_id) %>% 
  filter(mappability > 0.99)%>% 
    filter(!clone_id %in% c("0", "Pseudobulk"))

g044_cn_17_B <- plotCNprofile(cn_10kb[,cell_id := clone_id], chrstart = 25, chrend = 45,
                         tickwidth = 10,
              cellid = "B", 
              pointsize = 0.1,
              chrfilt = c("17"),
              SV = svs_md %>% filter(target_idx == "17_37721055") %>% as.data.frame(),
              ideogram = F,
              maxCN = 30,
              font_size = 7,
              line_size = 0.25,
              legend.position = "none",
              y_axis_trans = "squashy"
              )


g044_cn_2_B <- plotCNprofile(cn_10kb[,cell_id := clone_id], chrstart = 235, chrend = 245,
                         tickwidth = 5,
              cellid = "B", 
              pointsize = 0.1,
              chrfilt = c("2"),
              SV = svs_md  %>% filter(target_idx == "17_37721055") %>% as.data.frame(),
              ideogram = F,
              maxCN = 30,
              font_size = 7,
              line_size = 0.25,
              legend.position = "none",
              y_axis_trans = "squashy"
              )

g044_cn_17_E <- plotCNprofile(cn_10kb[,cell_id := clone_id], chrstart = 25, chrend = 45,
                         tickwidth = 10,
              cellid = "E", 
              pointsize = 0.1,
              chrfilt = c("17"),
              SV = svs_md  %>% filter(target_idx == "17_37721055") %>% as.data.frame(),
              ideogram = F,
              maxCN = 30,
              font_size = 7,
              line_size = 0.25,
              legend.position = "none",
              y_axis_trans = "squashy"
              )


g044_cn_2_E <- plotCNprofile(cn_10kb[,cell_id := clone_id], chrstart = 235, chrend = 245,
                         tickwidth = 5,
              cellid = "E", 
              pointsize = 0.1,
              chrfilt = c("2"),
              SV = svs_md  %>% filter(target_idx == "17_37721055") %>% as.data.frame(),
              ideogram = F,
              maxCN = 30,
              font_size = 7,
              line_size = 0.25,
              legend.position = "none",
              y_axis_trans = "squashy"
              )

recurrence_patient_044 <- recurrence %>% filter(str_detect(patient_id, "044")) %>% filter(row_number() == 1)
g044_tra <- svs %>% 
  filter(target_idx == "target_17_37721055") %>% 
  filter(bamtype == "uncollapsed") %>% 
  select(timepoint, alt_counts, alt_seq_reads, days) %>% 
  arrange(timepoint) %>% 
  ggplot(aes(x = days, y = alt_seq_reads)) +
  geom_line(size = 0.5, col = "grey70") +
  geom_point(size = 0.4) +
  geom_rect(data = recurrence_patient_044,
                                 aes(xmin = date, x = NULL, y = NULL, fill = NULL, col = NULL), 
                                 xmax = Inf, ymin = -Inf, ymax = Inf, 
                                 fill = "grey60", color = NA, alpha = .2) +
  scale_x_continuous(breaks = scales::breaks_width(365), guide = guide_axis(angle = 0)) +
  scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 4, base = 10), breaks = c(0, 10, 30, 100, 500)) +
  xlab("Days") +
  ylab("# reads")


plot_grid(NULL,
          plot_grid(g044_cn_2_B + removexaxis + ylab("B"), NULL, g044_cn_17_B + removexaxis + removeyaxis, ncol = 3, rel_widths = c(1,0.2,2)),
          NULL,
          plot_grid(g044_cn_2_E  + ylab("E"), NULL, g044_cn_17_E + removeyaxis, ncol = 3, rel_widths = c(1,0.2,2)),
          NULL,
          plot_grid(NULL,g044_tra,NULL, ncol = 3, rel_widths = c(0.1,1,0.1)), 
          ncol = 1, rel_heights = c(0.1, 1, -0.1,1.2,0.1,1.3)) %>% 
    save_plot(here("Figures/Fig4_mullerplots_erbb2.pdf"), 
          ., base_width = 0.25 * 2 * 89 * 0.039, base_height =2.1)
```


```{r}
plot_grid(plot_grid(brca1plot, gbrcarev),
          NULL,
          plot_grid(g107_3a, g107_3b, NULL, ncol = 3),
          NULL,
          plot_grid(g045_3a  + scale_y_continuous(breaks = c(0, 5,10), limits = c(0, NA)),  
                    g045_3b + scale_y_continuous(breaks = c(0, 5,10), limits = c(0, NA)), NULL, ncol = 3), ncol = 5,
          rel_widths = c(1, 0.1, 1, 0.1, 1)) %>% 
  save_plot(here("Figures/Fig4_gene_annotations.pdf"), ., base_width = config$default_plot_size$width, base_height = 1)

```


## Extended data figure

Create EDF7

```{r}

supp_fig_pts <- c("002", "004", "006", "014", "022", "037", "051", "075", "081", "083", "129")

plotlistsupp <- list()
i <- 1
for (pts in supp_fig_pts){
  print(pts)
  treeplot <- plot_grid(ggplot() + theme_cowplot(font_size = 7) +
                          removexaxis + removeyaxis + ggtitle(paste0(letters[i], "   ", pts)), 
                        mytreeplots[[pts]], NULL, NULL,
                        ncol = 2, 
                        rel_widths = c(1,1),
                        rel_heights = c(1.3,1))
  plotlistsupp[[pts]] <- cowplot::plot_grid(treeplot, 
                                           plotlist[[pts]], 
                                           ncol = 2,
                                           rel_widths = c(0.5, 1))
  i <- i + 1
}

```


```{r}

for_diversity <- svdf_clone %>% 
  left_join(recurrence %>% filter(recurrence_number == 1) %>% rename(patient = patient_id)) %>% 
  mutate(recurrence = days > date) %>% 
  group_by(patient, timepoint, days, recurrence) %>% 
  mutate(maxf = max(clone_frequency)) %>% 
  ungroup() %>% 
  filter(patient %in% config$patients_timeline)

for_diversity %>% 
  filter(tumorfraction_ClonalSV > 0 & maxf > 0) %>% 
  group_by(patient, timepoint, days, recurrence) %>% 
  mutate(f= clone_frequency / sum(clone_frequency)) %>% 
  summarise(ent = entropy::entropy(f), 
            nclones = sum(clone_frequency > 0),
            simp = sum(f^2)) %>% 
  ungroup() %>% 
  ggplot(aes(x = timepoint, y = nclones)) + 
  geom_point() +
  geom_line() +
  facet_wrap(~patient, scales = "free_x")

(gf <- for_diversity %>% 
  filter(maxf > 0) %>% 
  group_by(patient, timepoint, days, recurrence) %>% 
  mutate(f= clone_frequency / sum(clone_frequency)) %>% 
  ungroup() %>% 
  select(recurrence, patient, clone_id, timepoint, days, f) %>% 
  mutate(recurrence = ifelse(str_detect(patient, "004") & timepoint == 9, TRUE, recurrence)) %>% 
  group_by(patient) %>% 
  filter(timepoint == 1 | timepoint == max(timepoint)) %>% 
  ungroup() %>% 
  mutate(pt = str_remove(patient, "SPECTRUM-OV-")) %>% 
  ggplot(aes(x = recurrence, y = f, fill = clone_id)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = config$clonecols) +
  facet_wrap(~pt, nrow = 1, strip.position = "bottom") +
  scale_x_discrete(labels = c("Baseline", "Recurrence"),
                   guide = guide_axis(angle = 90)) +
  theme(strip.background = element_blank(),
        strip.placement = "outside") +
  xlab("") +
  labs(fill = "") +
  ylab("Clone frequency"))


(gnclone<- for_diversity %>% 
  filter(maxf > 0) %>% 
  group_by(patient, timepoint, days, recurrence) %>% 
  mutate(f= sum(clone_frequency > 0)) %>% 
  ungroup() %>% 
  select(recurrence, patient, clone_id, timepoint, days, f) %>% 
  mutate(recurrence = ifelse(str_detect(patient, "004") & timepoint == 9, TRUE, recurrence)) %>% 
  group_by(patient) %>% 
  filter(timepoint == 1 | timepoint == max(timepoint)) %>% 
  ungroup() %>% 
  mutate(pt = str_remove(patient, "SPECTRUM-OV-")) %>% 
  select(-clone_id) %>% 
  distinct() %>% 
  ggplot(aes(x = recurrence, y = f)) +
  geom_col(fill = "grey70") +
  facet_wrap(~pt, nrow = 1, strip.position = "bottom") +
  scale_x_discrete(labels = c("Baseline", "Recurrence"),
                   guide = guide_axis(angle = 90)) +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank()) +
  xlab("") +
  removexaxis +
  labs(fill = "") +
  ylab("# Clones\ndetected"))

diversity <- for_diversity %>% 
  filter(tumor_fraction > 0 & maxf > 0) %>% 
  #filter(patient != "SPECTRUM-OV-022") %>% 
  filter(nclones == 1) %>% 
  mutate(recurrence = ifelse(str_detect(patient, "004") & timepoint == 9, TRUE, recurrence)) %>% 
  group_by(patient, timepoint, days, recurrence) %>% 
  mutate(x = clone_frequency / sum(clone_frequency)) %>% 
  summarise(ent = entropy::entropy(x), 
            nclones = sum(clone_frequency > 0),
            simp = sum(clone_frequency^2)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  group_by(patient) %>% 
  filter(timepoint == 1 | timepoint == max(timepoint)) %>% 
  mutate(has_recurrence = any(recurrence)) %>% 
  filter(has_recurrence == TRUE) %>% 
  ungroup()

gdiv <- diversity %>% 
  select(patient, recurrence, ent) %>% 
  ungroup() %>% 
  pivot_wider(names_from = "recurrence", values_from = "ent", names_prefix = "rec_") %>% 
  ggpaired(., cond1 = "rec_FALSE", 
           cond2 = "rec_TRUE", 
           fill = "condition", 
           width = 0.3,
           point.size = 0.1,
           outlier.size = 0.1,
           line.size = 0.1) +
  scale_fill_manual(values = c("firebrick3", "deepskyblue4")) +
  stat_compare_means(method = "t.test", 
                     label.y.npc = 0.9,  
                     label = "p.format", 
                     size = 2,
                     #method.args = list(alternative = "less"),
                     label.x.npc = 0.5, 
                     paired = TRUE) +
  scale_x_discrete(labels = c("Baseline", "Recurrence"), 
                   guide = guide_axis(angle = 45)) +
  xlab("") +
  ylab("Shannon entropy") +
  theme_cowplot(font_size = 7, line_size = 0.25) +
  theme(legend.position = "none")
gdiv

gnclone_box <- diversity %>% 
  select(patient, recurrence, nclones) %>% 
  pivot_wider(names_from = "recurrence", values_from = "nclones", names_prefix = "n_") %>% 
  ggpaired(., cond1 = "n_FALSE", 
           cond2 = "n_TRUE", 
           fill = "condition", 
           width = 0.3,
           point.size = 0.1,
           outlier.size = 0.1,
           line.size = 0.1) +
    stat_compare_means(method = "t.test", 
                     label.y.npc = 0.9,  
                     size = 2,
                     #method.args = list(alternative = "less"),
                     label = "p.format", 
                     label.x.npc = 0.5, 
                     paired = TRUE) +
    scale_fill_manual(values = c("firebrick3", "deepskyblue4")) +
    scale_x_discrete(labels = c("Baseline", "Recurrence"), 
                   guide = guide_axis(angle = 45)) +
    xlab("") +
    ylab("Number of clones") +
    theme_cowplot(font_size = 7, line_size = 0.25) +
    theme(legend.position = "none")
gnclone_box
```


```{r}
for_leg <- plot_timeline("009", zerotimepoints = c(365, 1095), plotlist = T)
leg <- get_legend(for_leg$gclone)

gmuller_supp <- plot_grid(plot_grid(plotlist = plotlistsupp,
                   #labels = letters[1:10],
                   label_size = 8, ncol = 3),
                   leg,
                   ncol = 1,
                   rel_heights = c(1, 0.1))

gbottom1 <- cowplot:: plot_grid(gnclone, NULL, gf + ylab("Clone\nfrequency"), ncol = 1, rel_heights = c(0.35,-0.02, 1),
                    axis = "lr", align = "v")
gbottom <- plot_grid(gbottom1, gdiv, gnclone_box,
                     ncol = 3,
                     rel_widths = c(1, 0.3, 0.3), 
                     labels = c("l", "m", "n"),
                     label_size = 8)

gsupp <- plot_grid(gmuller_supp, 
                   gbottom, ncol = 1,
                   rel_heights = c(6,2))

gsupp %>% 
  save_plot(here("Figures/EDF7.pdf"), 
          ., base_width = 2 * 89 * 0.039, base_height =10)

```

```{r, fig.height=10, fig.width = 7}
gsupp
```


## Truncal SV vaf relative to recurrence

Here we want to pick out the dates and the difference (diff) when we detect ctDNA via SVs prior to the first clinical recurrence. This is just done manually.

```{r}
purity %>% 
  left_join(recurrence %>% filter(recurrence_number == 1) %>% rename(patient = patient_id)) %>% 
  select(patient, sample, timepoint, days, tumorfraction_ClonalSV, recurrence_number, date) %>% 
  mutate(diff = date - days) %>% 
  arrange(patient, timepoint)

c(76, #002
  184, #009
  117, #022
  509, #026
  336, #037
  109, #045
  363, #046
  269, #051
  173, #075
  314, #081
  224 #139
) %>% mean
```